---
title: "Actual vs. Statutory Effective Tariff Rates, 2025--2026"
author: "The Budget Lab at Yale"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
  word_document:
    toc: false
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(here)
library(patchwork)

here::i_am("R/etr_eval_report_v2.Rmd")
source(here("R", "utils.R"))
theme_set(theme_etr())

# ==============================================================================
# HS CHAPTER NAMES
# ==============================================================================

HS2_NAMES <- c(
  "01" = "Live Animals",
  "02" = "Meat",
  "03" = "Fish & Seafood",
  "04" = "Dairy, Eggs, Honey",
  "05" = "Other Animal Products",
  "06" = "Live Plants & Flowers",
  "07" = "Vegetables",
  "08" = "Fruits & Nuts",
  "09" = "Coffee, Tea, Spices",
  "10" = "Cereals",
  "11" = "Milling Products",
  "12" = "Oilseeds",
  "13" = "Gums & Resins",
  "14" = "Vegetable Plaiting Materials",
  "15" = "Fats & Oils",
  "16" = "Meat & Fish Preparations",
  "17" = "Sugars",
  "18" = "Cocoa",
  "19" = "Cereal & Flour Preparations",
  "20" = "Vegetable & Fruit Preparations",
  "21" = "Misc. Food Preparations",
  "22" = "Beverages",
  "23" = "Food Waste & Animal Feed",
  "24" = "Tobacco",
  "25" = "Salt, Sulfur, Earths, Stone",
  "26" = "Ores & Slag",
  "27" = "Mineral Fuels & Oils",
  "28" = "Inorganic Chemicals",
  "29" = "Organic Chemicals",
  "30" = "Pharmaceuticals",
  "31" = "Fertilizers",
  "32" = "Tanning & Dyeing Extracts",
  "33" = "Perfumery & Cosmetics",
  "34" = "Soap & Waxes",
  "35" = "Albuminoids & Glues",
  "36" = "Explosives",
  "37" = "Photographic Goods",
  "38" = "Misc. Chemical Products",
  "39" = "Plastics",
  "40" = "Rubber",
  "41" = "Raw Hides & Leather",
  "42" = "Leather Articles & Luggage",
  "43" = "Fur & Artificial Fur",
  "44" = "Wood",
  "45" = "Cork",
  "46" = "Straw & Basketware",
  "47" = "Wood Pulp",
  "48" = "Paper & Paperboard",
  "49" = "Printed Material",
  "50" = "Silk",
  "51" = "Wool",
  "52" = "Cotton",
  "53" = "Other Vegetable Fibers",
  "54" = "Man-Made Filaments",
  "55" = "Man-Made Staple Fibers",
  "56" = "Wadding & Nonwovens",
  "57" = "Carpets",
  "58" = "Special Woven Fabrics",
  "59" = "Coated Fabrics",
  "60" = "Knitted Fabrics",
  "61" = "Knitted Apparel",
  "62" = "Woven Apparel",
  "63" = "Other Textile Articles",
  "64" = "Footwear",
  "65" = "Headgear",
  "66" = "Umbrellas",
  "67" = "Prepared Feathers",
  "68" = "Stone & Cement Articles",
  "69" = "Ceramic Products",
  "70" = "Glass",
  "71" = "Precious Stones & Metals",
  "72" = "Iron & Steel",
  "73" = "Iron & Steel Articles",
  "74" = "Copper",
  "75" = "Nickel",
  "76" = "Aluminum",
  "78" = "Lead",
  "79" = "Zinc",
  "80" = "Tin",
  "81" = "Other Base Metals",
  "82" = "Metal Tools",
  "83" = "Metal Misc. Articles",
  "84" = "Machinery",
  "85" = "Electrical Equipment",
  "86" = "Railway Equipment",
  "87" = "Vehicles",
  "88" = "Aircraft",
  "89" = "Ships & Boats",
  "90" = "Optical & Medical Instruments",
  "91" = "Clocks & Watches",
  "92" = "Musical Instruments",
  "93" = "Arms & Ammunition",
  "94" = "Furniture & Lighting",
  "95" = "Toys & Games",
  "96" = "Misc. Manufactured Articles",
  "97" = "Art & Antiques",
  "98" = "Special Classification",
  "99" = "Special Import Provisions"
)

label_hs2 <- function(hs2) {

  name <- HS2_NAMES[hs2]
  ifelse(is.na(name), paste0("Ch ", hs2), paste0(hs2, ": ", name))
}

# ==============================================================================
# DATA LOADING
# ==============================================================================

# Actual ETR (customs duties / imports, monthly from Census/Treasury via Haver)
actual_etr <- load_actual_etr() %>%
  filter(date >= "2024-01-01") %>%
  mutate(actual_rate = effective_rate / 100)

# Daily statutory ETR (import-weighted, from tariff-rate-tracker)
daily_statutory <- load_daily_etr() %>%
  filter(date <= Sys.Date())

# Monthly statutory (first-of-month snapshot)
monthly_statutory <- collapse_statutory_monthly(daily_statutory)

# Census HS2 x country x month (from 01_pull_census_trade.R)
census_file <- here("data", "census_hs2_country_monthly.csv")
census_available <- file.exists(census_file)
if (census_available) {
  census <- load_census_trade() %>%
    mutate(
      cty_code = as.character(cty_code),
      partner_group = assign_partner_group(cty_code)
    )
}

# Statutory timeseries (product-level)
ts_file <- file.path(TRACKER_DIR, "data", "timeseries", "rate_timeseries.rds")
ts_available <- file.exists(ts_file)
if (ts_available) {
  ts <- load_timeseries()
}

# HTS10-level import weights (2024 annual, from tariff-rate-tracker)
# These are essential for correctly weighting product-level rates when
# aggregating to HS2 x country. Without them, unweighted means drastically
# overstate the ETR (e.g., 32% vs correct 4% in Feb 2025).
import_weights_file <- file.path(TRACKER_DIR, "config", "local_paths.yaml")
imports_hs10 <- NULL
if (file.exists(import_weights_file)) {
  local_paths <- yaml::read_yaml(import_weights_file)
  iw_path <- local_paths$import_weights
  # Resolve relative path from tracker dir
  if (!is.null(iw_path) && !startsWith(iw_path, "/")) {
    iw_path <- normalizePath(file.path(TRACKER_DIR, iw_path), mustWork = FALSE)
  }
  if (!is.null(iw_path) && file.exists(iw_path)) {
    imports_hs10 <- readRDS(iw_path) %>%
      group_by(hs10, cty_code) %>%
      summarize(imports = sum(imports, na.rm = TRUE), .groups = "drop") %>%
      filter(imports > 0) %>%
      mutate(cty_code = as.character(cty_code))
  }
}
has_hs10_weights <- !is.null(imports_hs10)

# ==============================================================================
# FIGURE/TABLE METADATA
# ==============================================================================

FIGURE_METADATA <- list(
  F1 = list(
    label    = "Figure 1",
    title    = "Actual vs. Statutory Effective Tariff Rate",
    subtitle = "Monthly actual (customs duties / imports) vs. daily statutory (import-weighted)",
    source   = "U.S. Treasury/Census via Haver Analytics; The Budget Lab Tariff Rate Tracker",
    notes    = "Actual rate is customs duties as a share of goods import value (monthly). Statutory rate is the import-weighted average of all HTS tariff schedules by calendar day.",
    y_label  = "Percent"
  ),
  F2 = list(
    label    = "Figure 2",
    title    = "Decomposing the Gap: Behavioral vs. Structural",
    subtitle = "Statutory-actual gap split into trade-weight shifts and residual implementation factors",
    source   = "The Budget Lab analysis; Census Bureau International Trade API",
    notes    = "Behavioral = difference between statutory ETR with 2024 weights and actual monthly weights. Residual = difference between monthly-weighted statutory ETR and Treasury actual collections.",
    y_label  = "Percentage Points"
  ),
  F3 = list(
    label    = "Figure 3",
    title    = "Monthly Gap by Major Trading Partner",
    subtitle = "Behavioral gap (2024 vs. actual monthly import weights) by country",
    source   = "The Budget Lab analysis; Census Bureau International Trade API",
    notes    = "Positive values = 2024 weights overstate the ETR for that partner (imports shifted toward lower-tariff products or declined).",
    y_label  = "Behavioral Gap (pp)"
  ),
  F4 = list(
    label    = "Figure 4",
    title    = "Monthly Gap by Sector",
    subtitle = "Behavioral gap (2024 vs. actual monthly import weights) by HS chapter",
    source   = "The Budget Lab analysis; Census Bureau International Trade API",
    notes    = "Positive values = 2024 weights overstate the ETR for that sector. Top sectors by 2024 import value shown.",
    y_label  = "Behavioral Gap (pp)"
  )
)

fig_meta <- function(id, field) {
  val <- FIGURE_METADATA[[id]][[field]]
  if (is.null(val)) "" else val
}
fig_caption <- function(id) {
  src <- fig_meta(id, "source")
  notes <- fig_meta(id, "notes")
  cap <- paste0("Source: ", src)
  if (nzchar(notes)) cap <- paste0(cap, "\nNote: ", notes)
  cap
}

# ==============================================================================
# COMPUTED OBJECTS
# ==============================================================================

comparison <- monthly_statutory %>%
  select(date, statutory_rate = weighted_etr) %>%
  full_join(
    actual_etr %>% select(date, actual_rate),
    by = "date"
  ) %>%
  filter(date >= "2025-01-01") %>%
  arrange(date) %>%
  mutate(
    gap_pp = (statutory_rate - actual_rate) * 100,
    statutory_pct = statutory_rate * 100,
    actual_pct = actual_rate * 100
  )

latest_comp <- comparison %>% filter(!is.na(actual_rate)) %>% filter(date == max(date))
latest_month <- format(latest_comp$date, "%B %Y")
latest_statutory <- sprintf("%.1f", latest_comp$statutory_pct)
latest_actual <- sprintf("%.1f", latest_comp$actual_pct)
latest_gap <- sprintf("%.1f", latest_comp$gap_pp)
```

## 1. Introduction and Headline Comparison

This report compares the **statutory effective tariff rate** (ETR)---the import-weighted average tariff implied by the Harmonized Tariff Schedule---against the **actual effective tariff rate**---customs duties collected divided by goods import value. As of `r latest_month`, the statutory ETR was `r latest_statutory`% while the actual collected rate was `r latest_actual`%, a gap of `r latest_gap` percentage points. Understanding why these measures diverge is critical for evaluating the fiscal and economic impact of the 2025--2026 tariff escalation.

```{r fig-actual-vs-statutory}
ggplot() +
  geom_step(data = daily_statutory,
            aes(x = date, y = weighted_etr * 100),
            color = color_statutory, linewidth = 0.7) +
  geom_point(data = actual_etr %>% filter(date >= "2025-01-01"),
             aes(x = date, y = actual_rate * 100),
             color = color_actual, size = 3, shape = 16) +
  geom_vline(data = POLICY_EVENTS %>% filter(date <= max(daily_statutory$date)),
             aes(xintercept = date), linetype = "dashed",
             color = tbl_gray, linewidth = 0.3) +
  geom_text(data = POLICY_EVENTS %>% filter(date <= max(daily_statutory$date)),
            aes(x = date, y = Inf, label = label),
            vjust = 1.3, hjust = -0.05, size = 2.3, color = tbl_gray) +
  annotate("text", x = as.Date("2025-01-15"), y = max(daily_statutory$weighted_etr * 100) * 0.95,
           label = "Statutory (daily)", color = color_statutory, hjust = 0, fontface = "bold") +
  annotate("text", x = as.Date("2025-01-15"), y = max(daily_statutory$weighted_etr * 100) * 0.88,
           label = "Actual (monthly)", color = color_actual, hjust = 0, fontface = "bold") +
  labs(
    title = fig_meta("F1", "title"),
    subtitle = fig_meta("F1", "subtitle"),
    x = NULL, y = fig_meta("F1", "y_label"),
    caption = fig_caption("F1")
  ) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  coord_cartesian(ylim = c(0, NA))
```

```{r tbl-monthly-comparison}
comparison %>%
  filter(!is.na(actual_rate)) %>%
  mutate(
    Month = format(date, "%b %Y"),
    `Statutory ETR (%)` = sprintf("%.2f", statutory_pct),
    `Actual ETR (%)` = sprintf("%.2f", actual_pct),
    `Gap (pp)` = sprintf("%.2f", gap_pp)
  ) %>%
  select(Month, `Statutory ETR (%)`, `Actual ETR (%)`, `Gap (pp)`) %>%
  kable(caption = "Monthly Statutory vs. Actual Effective Tariff Rates",
        align = c("l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

We collapse the daily statutory ETR to monthly using **first-of-month snapshots**, preserving the step-function nature of tariff policy. Census reports monthly import values without within-month timing, so day-weighting would add no precision.

---

## 2. Why Do Statutory and Actual Rates Diverge?

A growing literature has examined why predicted or statutory tariff rates diverge from actual collections. We draw on three key studies to frame the analysis.

### Azzimonti (Richmond Fed, August 2025)

[Azzimonti (2025)](https://www.richmondfed.org/publications/research/economic_brief/2025/eb_25-29) provides the most detailed decomposition to date, analyzing the 8.8 percentage point gap between the predicted average ETR (17.5%) and the actual ETR (8.7%) in May 2025. She identifies three channels:

- **Within-country product mix changes** (0.6 pp): Importers reduced volumes of highly tariffed goods or shifted purchase timing. A CFO Survey found 24.6% of respondents planned to "move up purchases."
- **Cross-country import share shifts** (2.7 pp): China imports fell 45% relative to 2024; Vietnam's share rose from 4% to over 6%.
- **Implementation frictions** (4.2 pp, dominant): Shipping timing delays (~30 days from China), deferred ACH payments (6-week lag), legacy exemptions not captured in policy data, customs system lags, and court challenges to IEEPA authority. Azzimonti finds that 31% of Canadian product lines with expected tariffs generated zero collections.

Her conclusion: the gap is primarily an **implementation** phenomenon, not a forecasting failure, and she projects convergence to 10--14% as enforcement catches up.

### Penn Wharton Budget Model (February 2026)

[Penn Wharton (2026)](https://budgetmodel.wharton.upenn.edu/issues/2026/2/23/effective-tariff-rates-and-revenues-updated-february-23-2026) distinguishes "pre-substitution" (holding import patterns fixed) from "post-substitution" (actual) ETRs. By December 2025, the actual ETR was 9.8%. They highlight two key behavioral responses:

- **USMCA surged to 88.2% utilization** by December 2025, up from ~35--50% before tariffs. Canadian and Mexican imports claiming USMCA faced ETRs under 5%.
- **Purchasing pattern changes** reduced tariff revenue by an estimated **$45.7 billion** relative to what a static import composition would have generated.

### Cui, Feng, Lu, and Tomlin (2025)

[Cui et al. (2025)](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=5110124) model the firm-level decision between stockpiling, sourcing diversification, and responsive pricing as tariff risk mitigation strategies. They find that **stockpiling dominates in the short run** when tariff timing is known, but **sourcing diversification is more sustainable** as tariff duration increases.

### Taxonomy of Channels

Drawing on this literature, we organize the sources of the gap into three categories:

| Category | Channel | Direction | Detectable at finer granularity? |
|----------|---------|-----------|----------------------------------|
| **Behavioral** | Trade diversion across countries | Widens gap | Yes: 2024 to 2025 weight shift |
| | Product mix shifts within countries | Widens gap | Yes: chapter-level reweighting |
| | Stockpiling / front-running | Widens gap (temporary) | Partially: import timing |
| **Structural** | USMCA & preference programs | Depends on model | Small residual (tracker uses utilization shares) |
| | Legacy exemptions & exclusions | Widens gap | No: persists at all levels |
| | De minimis threshold | Widens gap | No: not in HTS schedule |
| **Implementation** | Collection timing & ACH deferral | Widens gap (temporary) | No: aggregate revenue timing |
| | Customs system lags | Widens gap (temporary) | No: persists at all levels |
| | Court challenges (IEEPA) | Widens gap | No: systemic |

The critical test: if the gap is behavioral, it should **shrink** when we reweight from 2024 to actual 2025 trade patterns, because the gap comes from applying 2024 import composition to 2025 rates. If the gap is structural or implementation-driven, it should **persist** at all granularity levels.

---

## 3. Decomposing the Gap: Monthly

```{r monthly-decomp, include=FALSE}
# Compute decomposition for EVERY available 2025 month using monthly import weights
can_decompose <- census_available && ts_available && has_hs10_weights

if (can_decompose) {
  # --- Build import-weighted HS2 x country statutory rates ---
  # The timeseries has HTS10 x country rates. We must weight by HTS10-level
  # import values when aggregating to HS2 x country. An unweighted mean
  # drastically overstates rates (many low-import products with high tariffs).

  # 2024 annual HS2 x country Census weights (for reweighting across cells)
  weights_2024 <- census %>%
    filter(year == 2024) %>%
    group_by(hs2, cty_code) %>%
    summarize(imports_2024 = sum(con_val_mo, na.rm = TRUE), .groups = "drop")

  # Available 2025 months in census data
  available_months <- census %>%
    filter(year == 2025) %>%
    distinct(month) %>%
    arrange(month) %>%
    pull(month)

  # --- Aggregate decomposition by month ---
  decomp_monthly <- list()
  # --- By country x month ---
  country_monthly <- list()
  # --- By chapter x month ---
  chapter_monthly <- list()

  for (m in available_months) {
    query_date <- as.Date(sprintf("2025-%02d-01", m))
    month_label <- format(query_date, "%b %Y")

    # Statutory snapshot at first of month (HTS10 x country)
    snapshot <- ts %>%
      filter(valid_from <= query_date, valid_until >= query_date) %>%
      rename(hs10 = hts10, cty_code = country)

    # Join ALL import weights onto snapshot rates. Left join from imports
    # ensures zero-tariff products remain in the denominator (matching the
    # tracker's daily_overall.csv, which divides tariff revenue by total
    # imports, not just tariffed imports).
    snapshot_weighted <- imports_hs10 %>%
      left_join(
        snapshot %>% select(hs10, cty_code, total_rate),
        by = c("hs10", "cty_code")
      ) %>%
      mutate(
        total_rate = coalesce(total_rate, 0),
        hs2 = substr(hs10, 1, 2)
      )

    stat_hs2_cty <- snapshot_weighted %>%
      group_by(hs2, cty_code) %>%
      summarize(
        mean_rate = weighted.mean(total_rate, w = imports, na.rm = TRUE),
        imports_hs10 = sum(imports),
        .groups = "drop"
      )

    # Actual monthly imports (HS2 x country from Census)
    actual_imports <- census %>%
      filter(year == 2025, month == m) %>%
      select(hs2, cty_code, imports_actual = con_val_mo)

    # Merge: import-weighted statutory rates + 2024 Census weights + actual imports
    merged <- stat_hs2_cty %>%
      inner_join(weights_2024, by = c("hs2", "cty_code")) %>%
      left_join(actual_imports, by = c("hs2", "cty_code")) %>%
      mutate(imports_actual = coalesce(imports_actual, 0))

    # Coverage: what share of 2024 trade is represented after the inner_join?
    coverage_2024 <- sum(merged$imports_2024) / sum(weights_2024$imports_2024)

    # --- (A) Aggregate decomposition ---
    # Two-stage aggregation: HTS10 weights within each HS2 x country cell
    # (Stage 1, already done above), then HTS10-aggregated weights across
    # cells (Stage 2, here). Using imports_hs10 (sum of HTS10 weights per
    # cell) ensures etr_2024w reproduces the tracker's one-stage ETR exactly.
    # See docs/weighting_note.md for the derivation.
    #
    # etr_2024w: HS2 x country cells weighted by HTS10-aggregated 2024 imports
    # etr_actualw: same cells weighted by 2025 monthly Census imports
    # etr_hs10: HTS10-level reference (should match etr_2024w and daily_overall.csv)
    etr_hs10 <- weighted.mean(snapshot_weighted$total_rate,
                              w = snapshot_weighted$imports, na.rm = TRUE)
    etr_2024w <- weighted.mean(merged$mean_rate, w = merged$imports_hs10, na.rm = TRUE)

    merged_pos <- merged %>% filter(imports_actual > 0)
    etr_actualw <- if (nrow(merged_pos) > 0) {
      weighted.mean(merged_pos$mean_rate, w = merged_pos$imports_actual, na.rm = TRUE)
    } else NA_real_

    treasury_val <- actual_etr %>%
      filter(date == query_date) %>%
      pull(actual_rate)
    treasury_val <- if (length(treasury_val) == 1) treasury_val else NA_real_

    decomp_monthly[[month_label]] <- tibble(
      month = month_label,
      date = query_date,
      etr_statutory_hs10 = etr_hs10,
      etr_statutory_2024w = etr_2024w,
      etr_statutory_actualw = etr_actualw,
      etr_treasury = treasury_val,
      gap_total = etr_2024w - treasury_val,
      gap_behavioral = etr_2024w - etr_actualw,
      gap_residual = etr_actualw - treasury_val,
      coverage = coverage_2024
    )

    # --- (B) By country ---
    major_groups <- c("China", "Canada", "Mexico", "EU", "Japan", "S. Korea", "UK", "ROW")

    country_merged <- merged %>%
      mutate(partner_group = assign_partner_group(cty_code))

    cty_result <- country_merged %>%
      group_by(partner_group) %>%
      summarize(
        etr_2024w = weighted.mean(mean_rate, w = imports_hs10, na.rm = TRUE),
        etr_actualw = ifelse(
          sum(imports_actual) > 0,
          weighted.mean(mean_rate, w = imports_actual, na.rm = TRUE),
          NA_real_
        ),
        imports_hs10_B = sum(imports_hs10) / 1e9,
        imports_2024_B = sum(imports_2024) / 1e9,
        imports_actual_B = sum(imports_actual) / 1e9,
        .groups = "drop"
      ) %>%
      mutate(
        gap_pp = (etr_2024w - etr_actualw) * 100,
        month = month_label,
        date = query_date
      )
    country_monthly[[month_label]] <- cty_result

    # --- (C) By chapter ---
    ch_result <- merged %>%
      group_by(hs2) %>%
      summarize(
        etr_2024w = weighted.mean(mean_rate, w = imports_hs10, na.rm = TRUE),
        etr_actualw = ifelse(
          sum(imports_actual) > 0,
          weighted.mean(mean_rate, w = imports_actual, na.rm = TRUE),
          NA_real_
        ),
        imports_hs10_B = sum(imports_hs10) / 1e9,
        imports_2024_B = sum(imports_2024) / 1e9,
        imports_actual_B = sum(imports_actual) / 1e9,
        .groups = "drop"
      ) %>%
      mutate(
        gap_pp = (etr_2024w - etr_actualw) * 100,
        sector = label_hs2(hs2),
        month = month_label,
        date = query_date
      )
    chapter_monthly[[month_label]] <- ch_result
  }

  decomp_all <- bind_rows(decomp_monthly)
  country_all <- bind_rows(country_monthly)
  chapter_all <- bind_rows(chapter_monthly)

  # Prep for plotting
  decomp_plot <- decomp_all %>%
    filter(!is.na(etr_treasury)) %>%
    mutate(
      behavioral_pp = gap_behavioral * 100,
      residual_pp = gap_residual * 100,
      total_pp = gap_total * 100,
      behavioral_share = ifelse(total_pp != 0, behavioral_pp / total_pp * 100, NA_real_)
    )

  # --- Reconciliation: verify sub-decompositions aggregate back ---
  # Import-weighted sum of country-level behavioral gaps should equal aggregate
  reconcile <- decomp_all %>%
    select(date, month, agg_behavioral = gap_behavioral) %>%
    inner_join(
      country_all %>%
        group_by(date) %>%
        summarize(
          cty_behavioral = weighted.mean(
            etr_2024w - etr_actualw,
            w = imports_hs10_B,
            na.rm = TRUE
          ),
          .groups = "drop"
        ),
      by = "date"
    ) %>%
    mutate(
      discrepancy_pp = (agg_behavioral - cty_behavioral) * 100
    )
  max_discrep <- max(abs(reconcile$discrepancy_pp), na.rm = TRUE)
}
```

We compute the statutory ETR two ways for each month: (a) weighted by **2024 annual trade patterns** (matching the rate tracker's published ETR), and (b) weighted by **actual monthly imports** from Census data. The difference is the behavioral component---how much of the gap is explained by importers shifting *what* and *where* they buy. The residual, after subtracting both the behavioral component and the Treasury actual rate, captures structural and implementation factors.

Because HTS10-level data are available only annually, while Census reports monthly imports at HS2 x country, we use a two-stage aggregation: HTS10-level import-weighted rates within each HS2 x country cell, then cell totals across cells. The monthly-weight ETR is equivalent to constructing synthetic monthly HTS10 weights that match Census monthly cell totals while holding within-cell product composition fixed (see `docs/weighting_note.md` for the derivation). Within-cell product mix shifts are absorbed into the residual.

```{r fig-decomp-stacked, eval=can_decompose, fig.height=6}
decomp_long <- decomp_plot %>%
  select(date, month, behavioral_pp, residual_pp) %>%
  pivot_longer(c(behavioral_pp, residual_pp),
               names_to = "component", values_to = "pp") %>%
  mutate(component = factor(component,
    levels = c("residual_pp", "behavioral_pp"),
    labels = c("Structural / Implementation", "Behavioral (trade weight shift)")))

decomp_long %>%
  ggplot(aes(x = date, y = pp, fill = component)) +
  geom_col(position = "stack", width = 25) +
  geom_text(data = decomp_plot %>%
              mutate(label = ifelse(
                is.na(behavioral_share) | behavioral_share < 0 | behavioral_share > 100,
                sprintf("%.1f pp", total_pp),
                sprintf("%.1f pp\n(%.0f%% beh.)", total_pp, behavioral_share)
              )),
            aes(x = date, y = total_pp + 0.5, label = label),
            inherit.aes = FALSE, size = 2.5) +
  geom_point(data = decomp_plot,
             aes(x = date, y = total_pp),
             inherit.aes = FALSE, shape = 18, size = 3, color = "black") +
  scale_fill_manual(values = c(
    "Structural / Implementation" = color_actual,
    "Behavioral (trade weight shift)" = color_statutory
  )) +
  labs(
    title = fig_meta("F2", "title"),
    subtitle = fig_meta("F2", "subtitle"),
    x = NULL, y = fig_meta("F2", "y_label"),
    fill = NULL,
    caption = fig_caption("F2")
  ) +
  theme(legend.position = "bottom")
```

```{r tbl-decomp, eval=can_decompose}
decomp_plot %>%
  mutate(
    Month = month,
    `Statutory ETR, 2024 weights (%)` = sprintf("%.1f", etr_statutory_2024w * 100),
    `Statutory ETR, monthly weights (%)` = sprintf("%.1f", etr_statutory_actualw * 100),
    `Treasury actual ETR (%)` = sprintf("%.1f", etr_treasury * 100),
    `Total gap (pp)` = sprintf("%.1f", total_pp),
    `Behavioral (pp)` = sprintf("%.1f", behavioral_pp),
    `Residual (pp)` = sprintf("%.1f", residual_pp),
    `Behavioral share` = ifelse(is.na(behavioral_share), "—", sprintf("%.0f%%", behavioral_share))
  ) %>%
  select(Month, `Statutory ETR, 2024 weights (%)`, `Statutory ETR, monthly weights (%)`,
         `Treasury actual ETR (%)`, `Total gap (pp)`, `Behavioral (pp)`,
         `Residual (pp)`, `Behavioral share`) %>%
  kable(caption = "Monthly Decomposition of Statutory-Actual ETR Gap",
        align = c("l", rep("r", 7))) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r fig-gap-over-time, eval=FALSE, include=FALSE}
# Removed: line graph was unclear
```

```{r decomp-na, eval=!can_decompose, results='asis'}
cat("*Gap decomposition not available. Requires Census trade data, rate timeseries, and HTS10-level import weights (via local_paths.yaml).*")
```

---

## 4. Gap by Major Trading Partner

```{r reconciliation-note, eval=can_decompose, results='asis'}
if (max_discrep < 0.1) {
  cat(sprintf(
    "*Reconciliation check: the import-weighted sum of country-level behavioral gaps matches the aggregate to within %.2f pp across all months.*\n\n",
    max_discrep
  ))
} else {
  cat(sprintf(
    "**Note:** The import-weighted sum of country-level behavioral gaps deviates from the aggregate by up to %.1f pp, likely due to coverage gaps in the HS2 x country merge.\n\n",
    max_discrep
  ))
}
```

```{r fig-country-gap-monthly, eval=FALSE, include=FALSE}
# Removed: line graph was unclear
```

```{r fig-country-stacked, eval=can_decompose, fig.height=7}
major <- c("China", "Canada", "Mexico", "EU", "Japan", "S. Korea", "UK")

# Compute each country's import-weighted contribution to the overall gap
# behavioral contribution = country behavioral gap * (country hs10 imports / total hs10 imports)
country_contrib <- country_all %>%
  filter(partner_group %in% major, !is.na(gap_pp)) %>%
  group_by(date) %>%
  mutate(
    total_hs10_imports = sum(imports_hs10_B),
    import_share = imports_hs10_B / total_hs10_imports,
    behavioral_contrib_pp = (etr_2024w - etr_actualw) * import_share * 100
  ) %>%
  ungroup()

# Add residual as a separate "country" for stacking
residual_by_month <- decomp_plot %>%
  select(date, residual_pp) %>%
  mutate(partner_group = "Residual (structural / implementation)",
         behavioral_contrib_pp = residual_pp)

stacked_data <- bind_rows(
  country_contrib %>% select(date, partner_group, behavioral_contrib_pp),
  residual_by_month %>% select(date, partner_group, behavioral_contrib_pp)
) %>%
  mutate(partner_group = factor(partner_group,
    levels = c("Residual (structural / implementation)",
               "ROW", "UK", "S. Korea", "Japan", "EU", "Mexico", "Canada", "China")))

stacked_colors <- c(PARTNER_COLORS, "Residual (structural / implementation)" = color_actual)

stacked_data %>%
  ggplot(aes(x = date, y = behavioral_contrib_pp, fill = partner_group)) +
  geom_col(position = "stack", width = 25) +
  geom_point(data = decomp_plot,
             aes(x = date, y = total_pp),
             inherit.aes = FALSE, shape = 18, size = 3, color = "black") +
  geom_text(data = decomp_plot,
            aes(x = date, y = total_pp + 0.3,
                label = sprintf("%.1f pp", total_pp)),
            inherit.aes = FALSE, size = 2.5) +
  scale_fill_manual(values = stacked_colors) +
  labs(
    title = "Statutory-Actual Gap by Trading Partner",
    subtitle = "Country contributions to behavioral gap + aggregate residual, monthly",
    x = NULL, y = "Percentage Points", fill = NULL,
    caption = "Source: The Budget Lab analysis; Census Bureau International Trade API"
  ) +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2))
```

```{r fig-country-facet, eval=FALSE, include=FALSE}
# Removed: replaced by stacked bar above
```

```{r tbl-country-monthly, eval=can_decompose}
# Pivot to wide table: one row per partner, columns for each month
country_wide <- country_all %>%
  filter(partner_group %in% major) %>%
  mutate(gap_label = sprintf("%.1f", gap_pp)) %>%
  select(partner_group, month, gap_label) %>%
  pivot_wider(names_from = month, values_from = gap_label)

country_wide %>%
  rename(Partner = partner_group) %>%
  kable(caption = "Behavioral Gap by Partner (pp), Monthly",
        align = c("l", rep("r", ncol(country_wide) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r tbl-country-detail, eval=can_decompose}
# Summary table with import levels for latest available month
latest_month_date <- max(country_all$date)
latest_month_label <- format(latest_month_date, "%B %Y")

country_all %>%
  filter(date == latest_month_date, partner_group %in% major) %>%
  arrange(desc(imports_2024_B)) %>%
  mutate(
    Partner = partner_group,
    `2024 Imports ($B)` = sprintf("%.1f", imports_2024_B),
    `ETR, 2024 weights (%)` = sprintf("%.1f", etr_2024w * 100),
    `ETR, actual weights (%)` = sprintf("%.1f", etr_actualw * 100),
    `Behavioral gap (pp)` = sprintf("%.2f", gap_pp)
  ) %>%
  select(Partner, `2024 Imports ($B)`, `ETR, 2024 weights (%)`,
         `ETR, actual weights (%)`, `Behavioral gap (pp)`) %>%
  kable(caption = paste0("Country-Level Detail (", latest_month_label, ")"),
        align = c("l", rep("r", 4))) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## 5. Gap by Sector

```{r sector-prep, include=FALSE}
if (can_decompose) {
  # Top 20 sectors by 2024 import value
  top_sectors <- chapter_all %>%
    group_by(hs2, sector) %>%
    summarize(imports_2024_B = max(imports_2024_B), .groups = "drop") %>%
    arrange(desc(imports_2024_B)) %>%
    head(20) %>%
    pull(hs2)
}
```

```{r fig-sector-gap-heatmap, eval=can_decompose, fig.height=9, fig.width=11}
chapter_all %>%
  filter(hs2 %in% top_sectors, !is.na(gap_pp)) %>%
  mutate(sector = factor(sector, levels = rev(
    chapter_all %>%
      filter(hs2 %in% top_sectors) %>%
      group_by(sector) %>%
      summarize(imp = max(imports_2024_B), .groups = "drop") %>%
      arrange(imp) %>%
      pull(sector)
  ))) %>%
  ggplot(aes(x = date, y = sector, fill = gap_pp)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.1f", gap_pp)), size = 2.5) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, name = "Gap (pp)") +
  labs(
    title = fig_meta("F4", "title"),
    subtitle = fig_meta("F4", "subtitle"),
    x = NULL, y = NULL,
    caption = fig_caption("F4")
  ) +
  theme(axis.text.y = element_text(size = 9))
```

```{r fig-sector-lines, eval=FALSE, include=FALSE}
# Removed: line graph was unclear
```

```{r tbl-sector-monthly, eval=can_decompose}
# Wide table: sectors x months
sector_wide <- chapter_all %>%
  filter(hs2 %in% top_sectors, !is.na(gap_pp)) %>%
  mutate(gap_label = sprintf("%.1f", gap_pp)) %>%
  select(sector, month, gap_label) %>%
  pivot_wider(names_from = month, values_from = gap_label)

# Reorder by import value
sector_order <- chapter_all %>%
  filter(hs2 %in% top_sectors) %>%
  group_by(sector) %>%
  summarize(imp = max(imports_2024_B), .groups = "drop") %>%
  arrange(desc(imp)) %>%
  pull(sector)

sector_wide %>%
  mutate(sector = factor(sector, levels = sector_order)) %>%
  arrange(sector) %>%
  rename(Sector = sector) %>%
  kable(caption = "Behavioral Gap by Sector (pp), Monthly (Top 20 by Import Value)",
        align = c("l", rep("r", ncol(sector_wide) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE, font_size = 11)
```

```{r tbl-sector-detail, eval=can_decompose}
chapter_all %>%
  filter(date == latest_month_date, hs2 %in% top_sectors) %>%
  arrange(desc(imports_2024_B)) %>%
  mutate(
    Sector = sector,
    `2024 Imports ($B)` = sprintf("%.1f", imports_2024_B),
    `ETR, 2024 wts (%)` = sprintf("%.1f", etr_2024w * 100),
    `ETR, actual wts (%)` = sprintf("%.1f", etr_actualw * 100),
    `Gap (pp)` = sprintf("%.1f", gap_pp)
  ) %>%
  select(Sector, `2024 Imports ($B)`, `ETR, 2024 wts (%)`,
         `ETR, actual wts (%)`, `Gap (pp)`) %>%
  kable(caption = paste0("Sector-Level Detail (", latest_month_label, ")"),
        align = c("l", rep("r", 4))) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE, font_size = 11)
```

---

## 6. Summary

```{r summary-text, eval=can_decompose, results='asis'}
avg_behavioral <- mean(decomp_plot$behavioral_pp, na.rm = TRUE)
avg_residual <- mean(decomp_plot$residual_pp, na.rm = TRUE)
avg_total <- mean(decomp_plot$total_pp, na.rm = TRUE)
avg_share <- avg_behavioral / avg_total * 100
avg_coverage <- mean(decomp_plot$coverage, na.rm = TRUE) * 100

cat(sprintf(
  "Across the %d months with available data, the average total gap is **%.1f pp**, of which **%.1f pp** (%.0f%%) is behavioral and **%.1f pp** is structural/implementation. The behavioral component---the portion explained by importers shifting trade patterns away from 2024 baselines---remains a minority of the total divergence, consistent with Azzimonti's (2025) finding that implementation frictions dominate. The decomposition covers %.0f%% of 2024 import value on average (HS2 x country cells matched between the rate tracker and Census trade data).",
  nrow(decomp_plot), avg_total, avg_behavioral, avg_share, avg_residual, avg_coverage
))
```

---

```{r export-csvs, include=FALSE, eval=can_decompose}
# --- Export decomposition results as CSV for external analysis ---
output_dir <- here("output")

# (1) Overall gap by month
decomp_plot %>%
  transmute(
    date,
    month,
    etr_statutory_2024w_pct = etr_statutory_2024w * 100,
    etr_statutory_actualw_pct = etr_statutory_actualw * 100,
    etr_treasury_pct = etr_treasury * 100,
    total_gap_pp = total_pp,
    behavioral_pp,
    residual_pp,
    behavioral_share_pct = behavioral_share,
    coverage
  ) %>%
  write_csv(file.path(output_dir, "gap_decomposition.csv"))

# (2) By country x month
country_all %>%
  transmute(
    date,
    month,
    partner_group,
    etr_2024w_pct = etr_2024w * 100,
    etr_actualw_pct = etr_actualw * 100,
    behavioral_gap_pp = gap_pp,
    imports_hs10_B,
    imports_2024_B,
    imports_actual_B
  ) %>%
  write_csv(file.path(output_dir, "gap_by_country.csv"))

# (3) By sector x month
chapter_all %>%
  transmute(
    date,
    month,
    hs2,
    sector,
    etr_2024w_pct = etr_2024w * 100,
    etr_actualw_pct = etr_actualw * 100,
    behavioral_gap_pp = gap_pp,
    imports_hs10_B,
    imports_2024_B,
    imports_actual_B
  ) %>%
  write_csv(file.path(output_dir, "gap_by_chapter.csv"))
```

*Report generated on `r Sys.time()` using R `r R.version.string`*
